---
title: "ENVS 193DS Final"
author: "Natalie Parker"
format: pdf
format:
  pdf:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---

# Link GitHub repo

https://github.com/natty-nat/ENVS-193DS_spring-2026_final

# Set up

```{r}
#| label: libraries-and-data

library(tidyverse) # general use
library(janitor) # cleaning data frames
library(here) # file/folder organization
library(readxl) # reading .xlsx files
library(ggeffects)
# storing nest data as object "nest"
nest <- read_csv(here("data", "nest_data_final.csv"))
```

# Problem 1

## 1a.

In part 1 they used binary/ logistic regression.

In part 2 they used a Chi-Squared test for independence.

## 1b.

American Avocet microhabitat use is determined by the distance to the water's edge.

The shore birds American Avocet, Foster's Tern, and Black-necked Stilt, have preferences for certain habitats over others.

## 1c.

An additional variable that could effect American Avocet microhabitat choice could be their prey collected by looking in the soils. This is a continuous numeric variable (count/m^2^), and it might influence habitat use because with limited prey the carrying capacity is lower (less food), and vice versa.

# Problem 2

## 2a.

```{r}
#| label: cleaning-nest-data
# creating new object called "nests_clean" from "nest" data frame
nests_clean <- nest |> 
  # selecting columns of interest
  select(case_control, height_cm, ant_species) |> 
  # filtering to only include specific ant species
  filter(ant_species %in% c("AB", "RRB", "BBR")) |> 
  # making new column for scientifc names and adjusting to names to be full
  mutate(scientific_name = case_when(
    ant_species == "AB" ~ "Crematogaster sjostedti",
    ant_species == "RRB" ~ "Crematogaster mimosae",
    ant_species == "BBR" ~ "Crematogaster nigriceps"
  )) |> 
  # reordering the levels of ant species
  mutate(scientific_name = factor(scientific_name, levels = c(
    "Crematogaster sjostedti", 
    "Crematogaster mimosae", 
    "Crematogaster nigriceps"
  )))
# displaying the structure of "nests_clean"
str(nests_clean)
# showing 10 random rows!
nests_clean |> 
  sample_n(10)
```

## 2b.

The 1 and 0 in the column "case_control" tell the reader whether or not the tree contained a nest. The 1s are affermative for nest, where the 0s show that there was no nest.

## 2c.

Ant species is a categorical variable with 3 levels. This could influence the probability of nest occurrence depending on the type of ant species, and how aggressive they might be to the nesting birds. I found this information

Tree height is a continuous numerical value measured in cm. Tree height could influence the probability of nest occurrence because taller trees are more likely to have nests. I found this

## 2d.

| Model Number | Model Type | Formula | Model Description |
|----|----|----|----|
| Model 1 | Additive GLM | case_control \~ height_cm + scientific_name | Looking at the independent effect of tree height and ant species on nesting probability using logistic regression. |
| Model 2 | Interactive GLM | case_control \~ height_cm \* scientific_name | Looking at whether there is an effect of tree height on nesting probability changes with ant species present using logistic regression. |

: Table of models
